{\small} CUBEP3M: High Performance P3M N-body code

### 2012-04-17 Introduction.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

{\small PMFAST} (\cite{PMFAST}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for parallel computation.
This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

Since its first published version, {\small PMFAST} has evolved in many aspects. 
The first major improvement was to transform the volume decomposition in multi-node configurations 
from slabs to cubes. One of the problem with slabs is that they do not scale well to large runs: 
as the number of cells per dimension increases, the thickness of each slab shrinks rapidly,
until it reaches the hard limit of a single cell layer.  With this enhancement came a new name, {\small CUBEPM},
which soon after incorporated particle-particle (pp) interactions at the sub-grid level, and the code was renamed
{\small CUBEP3M}. It now includes a significant number of new features since MPT : the pp force
can be extended to an arbitrary range, the size of the redshift jump can be constrained for improved accuracy during the first time steps,
a runtime halofinder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features
{\bf (anything else?)}.
{\small CUBEP3M} has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
 \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)}

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It is structured as follow: XXXX

>>>>

{\small PMFAST} (\cite{PMFAST}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP}parallel computation.
This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

Since its first published version, {\small PMFAST} has evolved in many aspects. 
The first major improvement was to transform the volume decomposition in multi-node configurations 
from slabs to cubes. One of the problem with slabs is that they do not scale well to large runs: 
as the number of cells per dimension increases, the thickness of each slab shrinks rapidly,
until it reaches the hard limit of a single cell layer.  With this enhancement came a new name, {\small CUBEPM},
which soon after incorporated particle-particle (pp) interactions at the sub-grid level, and the code was renamed
{\small CUBEP3M}. It now includes a significant number of new features since MPT : the pp force
can be extended to an arbitrary range, the size of the redshift jump can be constrained for improved accuracy during the first time steps,
a runtime halofinder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features
{\bf (anything else?)}.

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
 \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
The paper is structured as follow: XXXX

### 2012-04-17 PID.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

\section{Other tools}

Particle ID, initial condition generator, projections, power spectrum code,  various expansions

>>>>

\subsection{Particle identification tags}
\label{PID}

A system of particle identification can be turned on, which basically allows to track each particle's trajectory
between checkpoints. Such a tool is useful to a number of applications, from reconstruction of halo merging history to {\bf (what else?)}
The particle tag system has been implemented as an array of double integers, {\tt PID}, 
and assigns a unique integer to each particle during the initialization stage. The location of the tag on the {\tt PID} array 
matches the location of the corresponding particle on the {\tt xv} array, hence it acts as if the latter array had an extra dimension.
The location change only when particles exist the local volume, in which case the tag is sent along with the particle in the {\tt pass\_particle} subroutine.
Deleted particles results in deleted flags, and the {\tt PID} array gets written to file at each particle checkpoint.
{\bf (Anything else to say here?)}

### 2012-04-17 PoissonSolver.tex
Completed first draft of Introduction.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

This section describes how Poisson's equation is solved at each time step of the main loop. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer extensively 
to section 2 of MPT.

The reason that motivates the double mesh configuration is to allow the force of gravity to be broken down into 
a long and a short range components. This two-mesh decomposition was first seen in Hydra \citep{hydra}, 
and is now used by many codes, including {\small PMFAST}, {\bf (TracPen2003)} and {\bf (Other codes?)}.
The long range force requires knowledge of the complete simulated volume, 
but can be computed on a coarser grid without degrading significantly the precision of the calculation.
Using a grid 64 times coarser limits the global communication time, plus contribute only to a fraction of the total memory requirement.
The short range force is computed on the finer mesh and is local in the sense that it only knows about the matter in the immediate surroundings.
It can thus be computed simultaneously on many computing units with {\small OPENMP} parallel loops, each {\small CPU} solving for a different sub-volume,
as long as the arrays are constructed such as to support parallel reading and writing\footnote{In practice, this is done by adding an additional dimension
to the relevant arrays, such that each {\small CPU} accesses a unique memory location.}. 
The particle-particle (pp) interactions are also local operations, and are computed inside the same {\small OPENMP} loops.

The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
  or a force  kernel $\omega_{F} ({\bf x})$.
It is a curl-free field, which allows us to relate the potential $\phi({\bf x})$to the source term via Poisson's equation: 
\begin{eqnarray}
\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
\label{eq:poisson}
\end{eqnarray}
$G$ being Newton's constant. We solve this equation in Fourier space, where we write
\begin{eqnarray}
 \tilde{\phi}({\bf k}) = \frac{4 \pi G \tilde{\rho}({\bf k})}{- k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
\label{eq:poissonFourier}
\end{eqnarray}
The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
Using the convolution theorem, we can write
\begin{eqnarray}
 \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}
\label{eq:poisson_solution_pot}
\end{eqnarray}
Although this approach is fast, it involves a finite differentiation which enhances the numerical noise.
We therefore opt for a force kernel, which is more accurate but has the inconvenient to require four extra Fourier transforms.
In this case, we must solve the convolution in three dimensions:
\begin{eqnarray}
 F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) {\bf  \omega}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
\label{eq:poisson_solution_force}
\end{eqnarray}
The differentiation does not affect the density since it only acts on un-prime variables,
and the force kernel is given by 
\begin{eqnarray}
{\bf  \omega}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
\end{eqnarray}

Following MPT, we split  the force kernel into two components, for the short and long range respectively, and 
match the overlapping regions with a polynomial. Namely, we have:
\begin{eqnarray}
{\bf \omega}_{s}(r) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
0 & \mbox{otherwise} 
\end{cases}
\end{eqnarray}
and
\begin{eqnarray}
{\bf \omega}_{l}(r) = \begin{cases} {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
 {\bf\omega}_{F}(r)  &\mbox{otherwise} 
\end{cases}
\end{eqnarray}
The vector $ {\bf \beta}(r)$ is related to the fourth order polynomial that is used in the potential case by
 $ {\bf \beta} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
 and we get
  \begin{eqnarray}
   {\bf \beta}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
  \end{eqnarray}
The long range force is subject to a correction that fixes {\bf (what exactly?)}.

 equation to match them, fudge factor,
corrected kernel, pp and extended pp.

>>>>

This section reviews how Poisson's equation is solved on a double-mesh configuration. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
to section 2 of MPT for more details. In cubep3m, the mass default assignment scheme are
a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
for the fine grid \citep{Hockney}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
for pairs closer to a tenth of a grid cell.

The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
  or a force  kernel $\omega_{F} ({\bf x})$.
Gravity fields are curl-free, which allows us to relate the potential $\phi({\bf x})$ to the source term via Poisson's equation: 
\begin{eqnarray}
\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
\label{eq:poisson}
\end{eqnarray}
$G$ being Newton's constant. We solve this equation in Fourier space, where we write
\begin{eqnarray}
 \tilde{\phi}({\bf k}) = \frac{4 \pi G \tilde{\rho}({\bf k})}{- k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
\label{eq:poissonFourier}
\end{eqnarray}
The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
Using the convolution theorem, we can write
\begin{eqnarray}
 \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}
\label{eq:poisson_solution_pot}
\end{eqnarray}
Although this approach is fast, it involves a finite differentiation which enhances the numerical noise.
We therefore opt for a force kernel, which is more accurate but has the inconvenient to require four extra Fourier transforms.
In this case, we must solve the convolution in three dimensions:
\begin{eqnarray}
 F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) {\bf  \omega}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
\label{eq:poisson_solution_force}
\end{eqnarray}
The differentiation does not affect the density since it only acts on un-prime variables,
and the force kernel is given by 
\begin{eqnarray}
{\bf  \omega}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
\end{eqnarray}

Following the spherically symmetric matching technique of MPT (section 2.1), 
we split  the force kernel into two components, for the short and long range respectively, and 
match the overlapping region with a polynomial. Namely, we have:
\begin{eqnarray}
{\bf \omega}_{s}({\bf r}) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
0 & \mbox{otherwise} 
\end{cases}
\end{eqnarray}
and
\begin{eqnarray}
{\bf \omega}_{l}(r) = \begin{cases} {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
 {\bf\omega}_{F}(r)  &\mbox{otherwise} 
\end{cases}
\end{eqnarray}
The vector $ {\bf \beta}(r)$ is related to the fourth order polynomial that is used in the potential case by
 $ {\bf \beta} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
 and we get
  \begin{eqnarray}
   {\bf \beta}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
  \end{eqnarray}

Because these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
both with the density and the kernel (see [Eq. 7-8] of MPT).
When matching the two  force kernels, it was realized that close to the cutoff region, the long range force is always on the low side, whereas 
the short range force is scattered across the theoretical $1/r^2$ value. These behaviours are purely features of the CIC and NGP interpolation scheme 
respectively. We identified a small range surrounding the cutoff length, in which we empirically adjust both kernels such as to improve 
the match. Namely, for $14 \le r \le 16$, ${\bf \omega}_{s}({\bf r}) \rightarrow {\bf \omega}_{s}({\bf r})*0.985$,
and for  $12 \le r \le 16$, ${\bf \omega}_{l}({\bf r}) \rightarrow {\bf \omega}_{l}({\bf r})*1.2$.
The two fudge factors were found by performing force measurements on two particles randomly placed in the volume.

In addition, the long range force is subject to a correction that fixes {\bf (what exactly fixes the LRCKCORR flag?)}.

{\bf (Anything else here?)}

### 2012-04-17 Structure.tex
Completed first draft of Introduction.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

\section{Presentation of the Code}
\label{sec:presentation}

An optimal N-body code is able to make efficient use of the available memory in order to maximize the size of the problem to solve,
and at the same time it must be fast for calculations to converge. In addition, communication time must be cut down to the strict minimum, 
and the load across all the available computing units must be as balanced as possible. 
In this section, we present the strategy adopted to address these challenges, and quickly walk through the code flow.
We refer the reader to future sections for detailed discussions on specific topics.
In its standard configuration, {\small CUBEP3M} is a {\small FORTRAN90} N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. The force of gravity from dark matter particles is decomposed into a sum of three distinct contribution,
which are constructed such as to optimize both speed and memory. The code has many extensions that departs from this basic scheme, and
we shall come back to these later.

\subsection{Memory foot-print and communication strategy}
\label{subsec:memory}
When designing a simulation geometry, one has to verify that the configuration fits in the system's memory.
The bulk of the memory foot-print is concentrated in a handful of large arrays that mostly concern the fine grid:
the coarse grid requires $4^3$ times less memory per node, hence has a minor contribution. 
In addition, many of these arrays are intermediate steps, hence it is possible to recycle the memory space required by the fine 
in the calculation of the coarse grid.  
Here are the largest  arrays used by the code:
\begin{enumerate}
\item{{\tt xv}, a phase-space array which stores the position and velocity of each particle} 
\item{{\tt ll}, a linked-list array that accelerate the access to particles in a local domain}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f}, a pair of three-dimensional array that stores 
the complete matter density for a given node in real and Fourier space respectively}
\item{{\tt force\_f}, an array that stores the force of gravity along the three Cartesian directions}
\item{{\tt kern\_f}, an array that stores the fine grid force kernel in the three directions}
\end{enumerate}

The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
Information is shared across multiple nodes with {\small MPI} communications. 
The short range force and the particle-particle interactions  are computed locally on a single processor,  
hence a node that has multiple cores can compute in parallel a series of such local contributions, or {\it tiles}, with the {\small OPENMP} technology.
The information exchange between cores that are part of the same motherboard is generally much faster than that across nodes,
hence it is generally a good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes.

\subsection{Code overview}
\label{subsec:overview}

The code flow is presented in Fig. \ref{fig:structure}.
Before entering the main loop, the code starts with an initialization stage, 
in which many declared variables are assigned default values,
the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
The phase-space array  is obtained from the output of the initial condition generator,
and the force kernels on both grids are also constructed from the specific geometry of the simulation.
For clarity, these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
although they are actually distinct calls in the code.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceed to a determination of the redshift jump by comparing the step size constraints from each
force components and from the cosmic expansion.
The latter is found by Taylor expanding  Friedmann's equation up to the third order in the scale factor.
Poisson's equation is then solved  in the {\tt particle\_mesh} subroutine,
which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney}.
If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and write them to file.

@@

\subsection{Two-level algorithm}
\label{subsec:particle_mesh}

The {\tt particle\_mesh} subroutine is conceptually identical to that of {\small PMFAST}, with the exception  that {\small CUBEP3M}
decomposes the volume into cubes (as opposed to slabs). The loop over tiles is thus performed in three dimension.
A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}, 
and we save the detailed description of the force calculation  in section \ref{sec:Poisson}.

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position
   call particle_pass
   call link_list
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f      
      call link_list_pp
      call force_pp
      call update_velocity_pp
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm.}
\label{fig:particle_mesh}
\end{figure}

force calculation,

>>>>

\section{Review of the Code Structure}
\label{sec:structure}

An optimal large scale N-body code must address many challenges: minimize the memory footprint to allow larger dynamical range,
minimize the passing of information across computing nodes, reduce and accelerate the memory accesses to the large scale arrays, 
make efficient use of high performance libraries to speed up standard calculations like Fourier transforms, just to name a few.
In the realm of parallel programming, high efficiency  can be assessed when a high load is balanced across all processors
most of the time. In this section, we present the general strategies adopted to address these challenges\footnote{ 
Many originate directly from MPT and were preserved in {\small CUBEP3M};
those will be briefly mentioned, and we shall refer the reader to the original PMFAST paper for greater details.}.
We start with a walkthrough the code flow, and briefly discuss some specific sections that depart from standard N-body codes,
while referring the reader to future sections for detailed discussions on selected topics.

As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. 
The code has extensions that departs from this basic scheme, and
we shall come back to these later, but for the moment, we adopt the 
standard configuration. 
The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing. In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles},
and for maximal efficiency, the number of tiles per node should be at least equal to the number of available processors.
Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes, as the information exchange between cores that are part of the same motherboard is generally much faster.
As mentioned in MPT, the computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned. The thickness of this surface must be larger than the cutoff length, and we find that a 24 cells deep buffer
is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}

%\subsection{Memory foot-print and communication strategy}
%\label{subsec:memory}

Because the coarse grid arrays require $4^3$ times less memory per node, 
the bulk of the foot-print is concentrated in a handful of fine grid arrays.
In addition, some of these are required for intermediate steps of the calculations only, 
hence it is possible to hide some of the coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in a local domain}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\end{enumerate}

%\subsection{Code overview}
%\label{subsec:overview}

The code flow is presented in Fig. \ref{fig:structure} and \ref{fig:particle_mesh}.
Before entering the main loop, the code starts with an initialization stage, 
in which many declared variables are assigned default values,
the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
The phase-space array  is obtained from the output of the initial conditions generator,
and the force kernels on both grids are constructed from the specific geometry of the simulation.
For clarity, all these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
although they are actually distinct calls in the code.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceeds to a determination of the redshift jump by comparing the step size constraints from each
force components and from the scale factor.
The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
and can accommodate constant or running equation of state of dark energy.
The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
The loop over tiles and the particle exchange are thus performed in three dimensions.
{\bf (Should I mention leapfrog here or anywhere else?)}
When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
and proceeds to the long range. This section of the code is also parallelized in many occasions, but, unfortunately, the current {\small MPI-FFTW}
do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
only the head processor is active. Other libraries such as {\small P3DFFT} or  Intel {\small MKL-$\alpha$} currently permit this extra level of parallelization,
and it is our plan to migrate to one of these in the near future.

@@

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position
   call link_list
   call particle_pass
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f
      if(pp = .true.) then       
         call link_list_pp
         call force_pp
         call update_velocity_pp
         if(extended_pp = .true.) then
            call link_list_pp_extended
            call force_pp_extended
            call update_velocity_pp_extended       
         endif
      endif
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm. We have included the section that concerns the extended pp force calculation to show that it follows the same basic logic. We mention here that the three linked list arrays are distinct entities, and their structure differ slightly.  }
\label{fig:particle_mesh}
\end{figure}

If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and writes them to file. Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time {\bf (is this correct?)}. 
The code exits the loop when it has reached the final redshift, it then wraps up the {\small FFTW} plans 
and clears the {\small MPI} communicators. We have collected those operations under the subroutine {\tt finalize} for concision.

{\bf (Anything else to say in this section?)}

### 2012-04-17 paper.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

\title[{\small} CUBEP3M: Closer to optimal N-body computations]{{\small} CUBEP3M: Closer to optimal N-body computations}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca} 
Hugh Merz$^{3}$ \thanks{E-mail: xxx} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: xxx}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Department of XXX\\
$^{4}$Department of XXX\\
$^{5}$Department of XXX}

@@

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{Accuracy} %pairwise and density force, 2-points stats.
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
%\input{MHD}
\input{PID}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3}{}
%\bibliographystyle{amsplain}

>>>>

\title[{\small} CUBEP3M: High Performance P3M N-body code]{{\small} CUBEP3M: High Performance P3M N-body code}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Hugh Merz$^{3}$ \thanks{E-mail: merz@sharcnet.ca} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Department of XXX\\
$^{4}$Department of XXX\\
$^{5}$Department of XXX}

@@

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{Accuracy} %pairwise and density force, 2-points stats.
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
\input{Extensions}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3}{}
%\bibliographystyle{amsplain}
